Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-08-09 Manuscript/manuscript.tex
Small updates

While the scaling factor of $f_q = 0.75$ for ions in water is well justified and shown
to improve model performance against scattering data \cite{kohagen14,kohagen16, Pluharova2014},
it is not clear if the same factor should be used for partial charges in molecules,
i.e. lipids in our case. Unlike the total charge of ions, atomic partial charges in
classical MD simulations are not experimental observables.
Instead they are based on theoretical calculations and the
derived values depend, for example, on the conformational
sampling of the molecules and description of solvent 
in the calculations \cite{RESP_paper,??}. 
%In cases where the partial charges have some physical meaning 
%(e.g. approximate electrostatic potential as in RESP~\cite{RESP_paper}),
Depending to the details, partial charge calculations may partially
incude some of the solvent electronic polarizability effects. Thus, we expect that
the correct ECC scaling factor, $f_q$, for molecular partial charges would lie
between $\epsilon _{el} ^{-1/2}$ (no electronic polarizability in calculation)
and $1$ (all electronic polarizability in calculation).

In this work, we empirically sample this parameter space
to find a atomistic MD simulation force field parameters for lipids that
accurately describes the lipid head group response to Na$^+$ and Ca$^{2+}$ concentrations
when compared with NMR data \cite{catte16}. This data can used to accurately asses
ion binding in PC bilayers, as discussed in section \ref{section:electrometer}.
We chose the Lipid14~\cite{dickson14} force field parameters as a starting point,
because their response to bound ions was apparently most realistic against NMR data
in recent work by NMRlipids project (see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and head group structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.
The ECC correction was applied to Lipid14 parameters by scaling partial charges of the headgroup,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding.
The hydrocarbon chain parameters are not modified, because they are
already highly optimized and give generally a good description for
hydrophobic part of lipid bilayers in various conditions
in most lipids models, including Lipid14 \cite{ollila16}.
In contrast, improvements in glycerol backbone and headgroup
parameters are required in all available lipid models~\cite{botan15}.

@@

Changes of headgroup order parameter as a function of added cationic surfactant
dihexadecyldimethylammonium bromide (C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) from
experiments~\cite{scherer89} and simulations are shown in
Fig. \ref{OrderParameterCHANGESnewMODELS}).
The head group order parameter response to the bound cation concentration
is approximately linear at least up to $\sim$0.3 mole fraction in experiments and
in both simulation models, original Lipid14 and ECClipids.

of the cation \cite{scherer89}.
The linearity is also observed in simulations with original Lipid14 and with
newly derived ECC-lipids models \todo{Yet to be shown.}.
Quantitative comparison, however, reveals that the response is overestimated in 
Lipid14 for both segments, while the ECC-lipids model gives a
good agreement for the change of $\alpha$ segment order parameter, and slightly
underestimates the $\beta$ segment order parameter change.
%Such a finding is not surprising, as the $\Delta S^\alpha/\Delta S^\beta$ ratio 
%of the response of Lipid14 model is found to be slightly larger than in 
%experiments (VALexp vs VALsim
%\todo{Add values of $S^\alpha/S^\beta$ response for sim and experiment
%  SAMULI: Maybe these could be put in the same figure with the order parameter
%  response and P-N vector angle change with surfactants.
%}
%), and this property is also kept by the newly derived Lipid14-based model, ECC-lipids. 
The present comparison of head group order parameter changes 
suggests that the observed overestimated response of all models to 
increasing CaCl$_2$ concentration in \cite{catte16} 
can be explained at least in part by a high sensitivity of the head group order parameter response
to the bound charge.

>>>>

While the scaling factor of $f_q = 0.75$ for ions in water is well justified and shown
to improve model performance against scattering data \cite{kohagen14,kohagen16, Pluharova2014},
it is not clear if the same factor should be used for partial charges in molecules,
i.e. lipids in our case. Unlike the total charge of ions, atomic partial charges in
classical MD simulations are not experimental observables.
Instead they are based on theoretical calculations and the
derived values depend, for example, on the conformational
sampling of the molecules and description of solvent 
in the calculations \cite{RESP_paper,??}. 
%In cases where the partial charges have some physical meaning 
%(e.g. approximate electrostatic potential as in RESP~\cite{RESP_paper}),
Depending to the details, partial charge calculations may partially
include some of the solvent electronic polarizability effects. Thus, we expect that
the correct ECC scaling factor, $f_q$, for molecular partial charges would lie
between $\epsilon _{el} ^{-1/2}$ (no electronic polarizability in calculation)
and $1$ (all electronic polarizability in calculation).

In this work, we empirically sample this parameter space
to find a atomistic MD simulation force field parameters for lipids that
accurately describes the lipid head group response to Na$^+$ and Ca$^{2+}$ concentrations
when compared with NMR data \cite{akutsu81,altenbach84}. This data can used to accurately asses
ion binding in PC bilayers, as discussed in Ref, \citenum{catte16} and in section \ref{section:electrometer}.
We chose the Lipid14~\cite{dickson14} force field parameters as a starting point,
because their response to bound ions was apparently most realistic against NMR data
in recent work by NMRlipids project (see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and head group structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.
The ECC correction was applied to Lipid14 parameters by scaling partial charges of the headgroup,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding.
The hydrocarbon chain parameters are not modified, because they are
already highly optimized and give generally a good description for
hydrophobic part of lipid bilayers in various conditions
in most lipids models, including Lipid14 \cite{ollila16}.
In contrast, improvements in glycerol backbone and headgroup
parameters are required in all available lipid models~\cite{botan15}.

@@

Changes of headgroup order parameter as a function of added cationic surfactant
dihexadecyldimethylammonium bromide (C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) from
experiments~\cite{scherer89} and simulations are shown in
Fig. \ref{OrderParameterCHANGESnewMODELS}).
The head group order parameter response to the bound cation concentration
is approximately linear at least up to $\sim$0.3 mole fraction in experiments and
in both simulation models, original Lipid14 and ECClipids \todo{Yet to be shown.}.

Quantitative comparison, however, reveals that the response is overestimated in 
Lipid14 for both segments, while the ECC-lipids model gives a
good agreement for the change of $\alpha$ segment order parameter, and slightly
underestimates the $\beta$ segment order parameter change.
%Such a finding is not surprising, as the $\Delta S^\alpha/\Delta S^\beta$ ratio 
%of the response of Lipid14 model is found to be slightly larger than in 
%experiments (VALexp vs VALsim
%\todo{Add values of $S^\alpha/S^\beta$ response for sim and experiment
%  SAMULI: Maybe these could be put in the same figure with the order parameter
%  response and P-N vector angle change with surfactants.
%}
%), and this property is also kept by the newly derived Lipid14-based model, ECC-lipids. 
The present comparison of head group order parameter changes 
suggests that the observed overestimated response of all models to 
increasing CaCl$_2$ concentration in \cite{catte16} 
can be explained at least in part by a high sensitivity of the head group order parameter response
to the bound charge.
